---
title: "Exercises Day 2"
author: "Richard Paquin Morel, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: FALSE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Load the `gapminder` dataset.

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
str(gapminder) #we need to convert some variables to factors

# gapminder$country = as.factor(gapminder$country)
# gapminder$continent = as.factor(gapminder$continent)
# str(gapminder)
```

In class exercise.
```{r}
#In Class Exercise

#1.
for (i in 1:nrow(gapminder)){
    gapminder[i,'log_gdpPercap'] = log(gapminder[i,'gdpPercap'])
    gapminder[i,'log_pop'] = log(gapminder[i,'pop'])
}

#2.
for (continent in unique(gapminder$continent)){
  mean_le = mean(gapminder$lifeExp[gapminder$continent == continent],na.rm = T)
  print(paste0(continent, ":",mean_le))
}

#3.
for (i in unique(gapminder$continent)) {
  print(paste0("Continent: ", i))
  for (j in unique(gapminder$year)) {
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i & gapminder$year == j], na.rm = T)
    print(paste0('      ', j, ": ", mean_le))
  }
}

#4.
for (i in unique(gapminder$continent)) {
  print(paste0("Continent: ", i))
  for (j in unique(gapminder$year)) {
    sd_le <- sd(gapminder$lifeExp[gapminder$continent == i & gapminder$year == j], na.rm = T)
    print(paste0('      ', j, ": ", sd_le))
  }
}

years = unique(gapminder$year)

#5. This one has errors -- need to define years
# sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year == x]))

#6.
i <-  1952 # define the interator
while (i < 1987) {
  sd_le <- sd(gapminder$lifeExp[gapminder$year == i],na.rm = T)
  print(paste0(i, ": ", sd_le))
  i <- i + 5 # increase the iterator by the interval between years
}

#7.
ini = 1987
while (ini <= 2002){
  sd_le = sd(gapminder$lifeExp[gapminder$year == ini], na.rm = T)
  print(paste0(ini, ": ", sd_le))
  ini = ini+1
}

#8.
for (i in years) {
  if (i >= 1987) {
    mean_pop <- mean(gapminder$lifeExp[gapminder$year == i])
    print(paste0(i, ": ", mean_pop))
  } else {
    print(paste("Sorry,", i ,"is less than 1987"))
  }
}

#9.
report_mean_sd <- 
  function(df, variable, country) {
    var <- df[[variable]][df$country == country] #df[,variableName] is the same as df[[variable]]
    m_le <- mean(var)
    sd_le <- sd(var)
    cat("Country:", country, 
        "\nMean Life Expectancy:", m_le,
        "\nSD Life Expectancy:", sd_le)
  }
report_mean_sd(gapminder, "lifeExp", "Bulgaria")

#10.
report_stats <- 
  function(df, variable, continent) {
    var <- df[[variable]][df$continent == continent] #df[,variableName] is the same as df[[variable]]
    meam_le <- mean(var)
    median_le = median(var)
    min_le = min(var)
    max_le = max(var)

    cat("Continent:", continent, 
        "\nMean Life Expectancy:", meam_le,
        "\nMedian Life Expectancy:", median_le,
        "\nMin Life Expectancy:", min_le,
        "\nMax Life Expectancy:", max_le)
  }
report_stats(gapminder, "lifeExp", "Asia")
```

## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{r}
if (sum(gapminder$year==2002)>0){
  print('There is more than one records from 2002 in the gapminder dataset.')
} else{
  print('No record from 2002 was found in the gapminder dataset.')
}

if (sum(gapminder$year==2012)>0){
  print('There is more than one records from 2012 in the gapminder dataset.')
} else{
  print('No record from 2012 was found in the gapminder dataset.')
}

#Archived Answer
# year<-2002
# if(any(gapminder$year == year)){
#    print(paste("Record(s) for the year",year,"found."))
# } else {
#   print(paste("No records for year",year))
# }

```


## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r}
for (country in unique(gapminder$country)){
  if (mean(gapminder$pop[gapminder$country==country]) < mean(gapminder$pop)){
    print(paste('Mean life expectancy in',country, 'is',
                mean(gapminder$lifeExp[gapminder$country==country])))
  } 
}

for (countinent in unique(gapminder$continent)){
  if (mean(gapminder$lifeExp[gapminder$continent==countinent]) < 50){
    print(paste('Mean Life Expectancy in',countinent ,'is less than 50'))
  } else if (mean(gapminder$lifeExp[gapminder$continent==countinent]) > 70){
    print(paste('Mean Life Expectancy in',countinent ,'is greater than 70'))
  } else{
    print(paste('Mean Life Expectancy in',countinent ,'is between 50 and 70'))
  }
}

#-------------------------------------------------------------------------------

### Archived Answer 1
overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   country_mean <- mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop

#Archived Answer 2
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
}
```

## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{r}
report_colname_class = 
  function(df) {
    for (col in 1:ncol(df)) {
      print(paste('Data type of column',colnames(df)[col], 'is:', class(df[,col])))
    }
  }

report_colname_class(gapminder)

#-------------------------------------

##Archived Answer
data_frame_info <- function(df) {
  cols <- names(df)
  for (i in cols) {
    print(paste0(i, ": ", mode(df[, i])))
  }
}
data_frame_info(gapminder)
```



Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r}
vector_stats_printer = 
  function(x,include_median = FALSE) {
    cat(
        "Mean value:", mean(x,na.rm = T),
        "\nStandard Deviation:", sd(x,na.rm = T)
        )
    if (include_median){
      cat("\nMedian value:", median(x,na.rm = T)
        )
  }
}
vector_stats_printer(1:30,T)


#------------------------
#Archived Answwer
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}

le <- gapminder$lifeExp
vector_info(le, include_median = F)
vector_info(le, include_median = T)
```


## Analyzing the relationship 

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)
```{r}
plot(x=log(gapminder$gdpPercap),y=log(gapminder$lifeExp))

for (year in unique(gapminder$year)){
  plot(x=log(gapminder$gdpPercap[gapminder$year==year]),y=log(gapminder$lifeExp[gapminder$year==year]))
}
```


2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.
```{r}
for (continent in unique(gapminder$continent)){
  plot(x=log(gapminder$gdpPercap[gapminder$continent==continent]),
       y=log(gapminder$lifeExp[gapminder$continent==continent]))
}

```

